---
title: "MB5370"
author: "Darius HUISMAN"
output: html_document
date: "2026-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 1

```{r}
source("../MB5370/MB5370_W1.R")
```


```{r Workshop 1}
#----------------------------------------------#
#MB5370: Introduction to programming
#Darius HUISMAN
# May 2026

#Workspace: 
# ~/Desktop/MB5370/MB5370_W1.R
#----------------------------------------------#

#----------------------------------------------#
# Workshop 01. Introduction ####

# Getting started ####
# This section introduces us to R by using it as a simple calculator.

# Variables and Assignment ####
# This section focuses on understanding how data is stored in R and why that matters.

age <- 25
first_name <- 'Bill'

Age <- 41 # cases matter

  # Exercise ####
a <- 15 + 25.1 + 20.25
b <- 26
c <- a + b

# Functions ####

years_old <- 25.7
round(years_old) # rounds up
floor(years_old) # rounds down

years_old <- 25.765
round (years_old, 2) # comma after the object to specify argument

?round # go to help

args(round) # use args in the Console

  # Exercise 
my_age <- 26
my_name <- 'Darius'

?paste # go to help

paste(my_name, "is", my_age, "years old.")

# Debugging ####

# Misconceptions
# variables in programs do not work the same way as they do in spreadsheets
grade <- 55
total <- grade + 10
print (total)
grade <- 90
print (total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)
total <- grade + 10
print (total) # executed in the way it was defined

# What will this code do?
p <- 2
z <- 5
out <- p * z  # What should the value of out be?
print (out) # What is the value of out? Is it the same as expected?

# Testing
x <- 1
is.character(x)
is.numeric (x)

  # Exercise 
#my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "cuatro" # missing a ",", causing an error
#             "cinco")
#print (my_quis) # syntax error, should be "my_quiz"
#str(my_quiz)
#len(my_quiz) # wrong function / function does not exist

# errors fixed
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro", # corrected
             "cinco")
print (my_quiz) # corrected 
str(my_quiz)
length(my_quiz) # corrected

# Data types ####

  #Exercise
#numeric
my_numeric_variable = 4.2 
class(my_numeric_variable)

#character
my_name <- 'Darius'
class(my_name)

#integer
int <- 1L
class(int)

#logical
t <- TRUE
class(t)

# Data structure ####

# Vector
y <- c(1, 2, 3)
z <- c("Sarah", "Tracy", "Jon")

  # Exrecise
class(x)
class(y)
## both are numeric

# List
x <- list(1, "a", TRUE)
x
x[[2]]

# Data frame and tibbles
my_data_frame <- data_frame(no = c(1,2,3), name = c("Tracey", "John", "Pete"), "True or False" = c(TRUE, FALSE, TRUE))
my_data_frame
str (my_data_frame)

my_data_frame$no = as.factor(my_data_frame$no)
str (my_data_frame)

# Packages and Libraries ####

# install.packages('tidyverse') # download and install
library(tidyverse) # load into current workspace

# install.packages ('ggplot2') # comment out so it doesn’t run again

# install.packages("ggplot","tidyr", "earthtones","redlistr","dplyr" ,"terra")

library (ggplot2)
?ggplot2

# Coding best practices ####

# load packages
library(ggplot2)
library(tidyr)
library(tidyverse)

  # Class discussion
    # 1. Read the .csv dataset into R using the 'read_csv' function
    # 2. Check the data file donwloaded (make sure the file has been read correctly and spot potential data that could cause problem)
    # 3. Analyse the abundance of each species
    # 4. PLot the species abundance of the species of you want 
```


# Workshop 2
This workshop aims to develop and master our skills in data visualisation using ggplot2

## 2.16 Load packages and data

```{r Packages}
install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
```

```{r mpg data}
mpg
```

## 2.17 Create your first ggplot

```{r first ggplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.18 Understand the ‘grammar of graphics’
### 2.18.1 Graphing template

```{r Graphing template}
#ggplot(data = mpg)
```

### 2.18.2 Aesthetic mappings

```{r Aesthetic mappings}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))

#change point size by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
#> Warning: Using size for a discrete variable is not advised.

#change point transparency by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

#change point shape by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

# all points blue
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")

#distinct displ data < 5
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

## 2.19 Troubleshooting

```{r Troubleshooting}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

## 2.20 Facet and panel plots

```{r Facet and panel plots}
# facet_wrap
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# facet_grid
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)

#using a .
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
# it does not make a plot for each hwy (compared to previous plot)
```
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 
```{r}
?facet_wrap
```
nrow, ncol	= number of rows and columns you want in the grid

## 2.21 Fitting simple lines

```{r Fitting simple lines}
#display data as point
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

#display same data as smooth line through the points
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

#with different dashes for drv
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))

#use group() argument
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

#Change the color of each line based on drv value
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )

#plot multiple geom() at the same time
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

#more efficent code
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

#style the points by themself
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()

#use a filter (class = "subcompact") to select a subset of the data and plot only that subset
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## Exercise
1.What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
  geom_line()
  geom_boxplot()
  geomp_histogram()
  geom_area()

2.Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions. Will these two graphs look different? Why/why not?
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

the two plots will provide with the same output. They will display the data with both points and smooth information at the smae time

```{r Exercise}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

## 2.22 Transformations and Stats

### 2.22.1 Plotting statistics

```{r Plotting statistics}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

#recreate the previous plot using stat_count() instead of geom_bar()
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

### 2.22.2 Overriding defaults in ggplot2

```{r Overriding defaults in ggplot2}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

#override a default mapping from transformed variables to aesthetics
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```

### 2.22.3 Plotting statistical details

```{r Plotting statistical details}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## 2.23 Aesthetic adjustments

```{r Aesthetic adjustments}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))

#colour by another variable like clarity
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

#use position = "identity"
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) +
  geom_bar(fill = NA, position = "identity")

#position = "fill" 
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

#position = "dodge"
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

#position = "jitter"
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```


# Workshop 4

The Developer's Toolbox & AI

# 4.4 chattr (Conversational Workspace Engine)

```{r chattr}
#install.packages("ellmer")
# install.packages("chattr")
```
# Exercise: Prompting for Context
```{r Exercise: Prompting for Context}
library(ggplot2)

data(iris)

ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point()
```

# 4.3 Conditional Logic - filtering your data

Base R: if and else
```{r if_else}
# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```


Tidyverse Alternative: Vectorised Conditionals (dplyr)

if_else()

```{r if_else tidyverse}
# Load tidyverse (if not already loaded)
library(tidyverse)

# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))
```


what mutate() does?
  The mutate() function, from the dplyr package, is used to add new variables (columns) to a data frame or modify existing ones. It always   operates on a data frame and returns a new data frame with the added or modified columns, while preserving existing columns.

what abou the  %>% symbol?
  The %>% symbol is known as the pipe operator. It comes from the magrittr package, which is loaded as part of the tidyverse.
  Its primary purpose is to pass the result of the expression on its left-hand side as the first argument to the function on its right-hand side. This allows you to chain multiple operations together in a more readable and sequential manner, making your code flow from left to right, rather than nesting functions inside each other.


case_when()

```{r case_when}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
```

task
```{r task case_when}
library(dplyr)

marine_stations <- tibble(
  salinity = c(35, 28, 32, 12)
) %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine",
      TRUE ~ NA_character_ # Optional: handles any other cases not explicitly covered
    )
  )

print(marine_stations)
```

# 4.4 Automation — Iterating Efficiently

Base R: for-loops
```{r for-loops}
# Classic loop anatomy
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}

# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```

Tidyverse Alternative: Functional Programming (purrr)
  Comparison Example: Calculating Square Roots of Site Vectors
  Using a standard for-loop:
```{r std for-loop}
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])
}
```

  Using the purrr equivalent:
```{r purr equivalent}
library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)

```


Iterating Across Complex Groups
```{r iterate}
# Iterate a summary function over split lists of data
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))
```

Task using chattr: 
```{r task purr}
library(purrr) # For map_dbl()

# List of fish count data
fish_counts_list <- list(
  site_a = c(10, 12, 8, 15, 11),
  site_b = c(25, 20, 22, 18),
  site_c = c(5, 7, 6)
)

# Using a for-loop
mean_counts_for_loop <- vector("numeric", length = length(fish_counts_list))
names(mean_counts_for_loop) <- names(fish_counts_list)

for (i in seq_along(fish_counts_list)) {
  mean_counts_for_loop[i] <- mean(fish_counts_list[[i]])
}

print("Mean counts (for-loop):")
print(mean_counts_for_loop)

# Using purrr's map_dbl()
mean_counts_map_dbl <- map_dbl(fish_counts_list, mean)

print("Mean counts (map_dbl):")
print(mean_counts_map_dbl)
```

# 4.5 Writing custom functions

Function Construction Blueprint
```{r}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)
```

Student Exercise 4: Custom Function Construction Task
```{r Ex custom fctn}
#create a function to convert temperature in C° to F°
convert_temp_c_to_f <- function(temp_C){
  #make a safety check to prevent temperatures below absolute zero
  if (temp_C < -273.15){
    stop("Entered value is below absolute zero (-273.15 degrees Celsius)")
  }
  temp_F <- (temp_C * 9/5) + 32
  return(temp_F)
}
#test function
convert_temp_c_to_f(temp_C = 25)
convert_temp_c_to_f(temp_C = -300)
```

# 4.6 Advanced Scripting Extension Challenges

Challenge 1: Nested Multi-Vector Iteration
```{r C1 Nested Multi-Vector Iteration}
survey_counts <- c(15, 24, 8, 42)
species_specific_scaling_factors <- c(1.2, 0.8, 2.5, 1.1)

purrr::map2_dbl(survey_counts,species_specific_scaling_factors, `*`)
```

Challenge 2: The Robust Data Wrangling Pipeline Function
```{r C2 The Robust Data Wrangling Pipeline Function}
#Create a custom function
clean_and_classify_survey <- function(df, col_name, threshold_parameter){
  #Safely check if the specified column name exists
  if (!col_name %in% names(df)){
    stop("Entered column ", col_name, " not found.")
  }
  #filter out NA values and mutate based on threshold
  cleaned_df <- df %>%
                  filter_out(is.na(!!sym(col_name))) %>%
                          mutate(status_classification = case_when(
                            col_name < threshold_parameter ~ "Fail",
                            col_name >= threshold_parameter ~ "Pass"
                          ))
  return(cleaned_df)
}

#tests
#1
test1 <- tibble(Names = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J"),
                Grade = c(56, 78, 42, 90, 87, 38, 67, 75, 2, 52))
clean_and_classify_survey(test1, "Grade", 50)

#2 (NA values)
test2 <- tibble(Names = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J"),
                Grade = c(56, 78, 42, 90, NA, 38, 67, 75, 2, NA))
clean_and_classify_survey(test2, "Grade", 50)


```


```{r}


```

