---
title: "MB5370"
author: "Darius HUISMAN"
output: html_document
date: "2026-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 1

```{r}
source("../MB5370/MB5370_W1.R")
```


```{r Workshop 1}
#----------------------------------------------#
#MB5370: Introduction to programming
#Darius HUISMAN
# May 2026

#Workspace: 
# ~/Desktop/MB5370/MB5370_W1.R
#----------------------------------------------#

#----------------------------------------------#
# Workshop 01. Introduction ####

# Getting started ####
# This section introduces us to R by using it as a simple calculator.

# Variables and Assignment ####
# This section focuses on understanding how data is stored in R and why that matters.

age <- 25
first_name <- 'Bill'

Age <- 41 # cases matter

  # Exercise ####
a <- 15 + 25.1 + 20.25
b <- 26
c <- a + b

# Functions ####

years_old <- 25.7
round(years_old) # rounds up
floor(years_old) # rounds down

years_old <- 25.765
round (years_old, 2) # comma after the object to specify argument

?round # go to help

args(round) # use args in the Console

  # Exercise 
my_age <- 26
my_name <- 'Darius'

?paste # go to help

paste(my_name, "is", my_age, "years old.")

# Debugging ####

# Misconceptions
# variables in programs do not work the same way as they do in spreadsheets
grade <- 55
total <- grade + 10
print (total)
grade <- 90
print (total) # value of total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65)
total <- grade + 10
print (total) # executed in the way it was defined

# What will this code do?
p <- 2
z <- 5
out <- p * z  # What should the value of out be?
print (out) # What is the value of out? Is it the same as expected?

# Testing
x <- 1
is.character(x)
is.numeric (x)

  # Exercise 
#my_quiz <- c("uno",
#             "dos",
#             "tres",
#             "cuatro" # missing a ",", causing an error
#             "cinco")
#print (my_quis) # syntax error, should be "my_quiz"
#str(my_quiz)
#len(my_quiz) # wrong function / function does not exist

# errors fixed
my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro", # corrected
             "cinco")
print (my_quiz) # corrected 
str(my_quiz)
length(my_quiz) # corrected

# Data types ####

  #Exercise
#numeric
my_numeric_variable = 4.2 
class(my_numeric_variable)

#character
my_name <- 'Darius'
class(my_name)

#integer
int <- 1L
class(int)

#logical
t <- TRUE
class(t)

# Data structure ####

# Vector
y <- c(1, 2, 3)
z <- c("Sarah", "Tracy", "Jon")

  # Exrecise
class(x)
class(y)
## both are numeric

# List
x <- list(1, "a", TRUE)
x
x[[2]]

# Data frame and tibbles
my_data_frame <- data_frame(no = c(1,2,3), name = c("Tracey", "John", "Pete"), "True or False" = c(TRUE, FALSE, TRUE))
my_data_frame
str (my_data_frame)

my_data_frame$no = as.factor(my_data_frame$no)
str (my_data_frame)

# Packages and Libraries ####

# install.packages('tidyverse') # download and install
library(tidyverse) # load into current workspace

# install.packages ('ggplot2') # comment out so it doesn’t run again

# install.packages("ggplot","tidyr", "earthtones","redlistr","dplyr" ,"terra")

library (ggplot2)
?ggplot2

# Coding best practices ####

# load packages
library(ggplot2)
library(tidyr)
library(tidyverse)

  # Class discussion
    # 1. Read the .csv dataset into R using the 'read_csv' function
    # 2. Check the data file donwloaded (make sure the file has been read correctly and spot potential data that could cause problem)
    # 3. Analyse the abundance of each species
    # 4. PLot the species abundance of the species of you want 
```


# Workshop 2
This workshop aims to develop and master our skills in data visualisation using ggplot2

## 2.16 Load packages and data

```{r Packages}
install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
```

```{r mpg data}
mpg
```

## 2.17 Create your first ggplot

```{r first ggplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.18 Understand the ‘grammar of graphics’
### 2.18.1 Graphing template

```{r Graphing template}
#ggplot(data = mpg)
```

### 2.18.2 Aesthetic mappings

```{r Aesthetic mappings}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))

#change point size by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
#> Warning: Using size for a discrete variable is not advised.

#change point transparency by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

#change point shape by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))

# all points blue
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")

#distinct displ data < 5
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

## 2.19 Troubleshooting

```{r Troubleshooting}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

## 2.20 Facet and panel plots

```{r Facet and panel plots}
# facet_wrap
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# facet_grid
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)

#using a .
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
# it does not make a plot for each hwy (compared to previous plot)
```
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 
```{r}
?facet_wrap
```
nrow, ncol	= number of rows and columns you want in the grid

## 2.21 Fitting simple lines

```{r Fitting simple lines}
#display data as point
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

#display same data as smooth line through the points
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

#with different dashes for drv
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))

#use group() argument
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

#Change the color of each line based on drv value
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )

#plot multiple geom() at the same time
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

#more efficent code
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

#style the points by themself
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()

#use a filter (class = "subcompact") to select a subset of the data and plot only that subset
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## Exercise
1.What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
  geom_line()
  geom_boxplot()
  geomp_histogram()
  geom_area()

2.Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions. Will these two graphs look different? Why/why not?
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

the two plots will provide with the same output. They will display the data with both points and smooth information at the smae time

```{r Exercise}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

## 2.22 Transformations and Stats

### 2.22.1 Plotting statistics

```{r Plotting statistics}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))

#recreate the previous plot using stat_count() instead of geom_bar()
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

### 2.22.2 Overriding defaults in ggplot2

```{r Overriding defaults in ggplot2}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

#override a default mapping from transformed variables to aesthetics
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```

### 2.22.3 Plotting statistical details

```{r Plotting statistical details}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## 2.23 Aesthetic adjustments

```{r Aesthetic adjustments}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))

#colour by another variable like clarity
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

#use position = "identity"
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) +
  geom_bar(fill = NA, position = "identity")

#position = "fill" 
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

#position = "dodge"
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

#position = "jitter"
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```























